---
title: "Descriptive Statistics - NAFLD vs. NASH Cohort"
author: "Catherine Cioffi"
date: "November 20, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction:

The goal of this analysis was to describe the metabolomic pathways associated with NAFLD disease progression. The following will generate descriptive Tables 1a and 1b of the characteristics of the comparison groups for this analysis. The first comparison was based on presence of lobular or portal inflammation, versus no inflammation. The second comparison was based on a NAFLD Activity SCore (NAS) greater than or equal to 3, versus less than 3.

First we will read in the phenotype table from computer hard-drive:

```{r}
library(readxl)
data <- read.csv("C:/Users/cecioff/Desktop/metabolomics/_CHEAR/NASH_panel/phenotype_data.csv", header=TRUE, sep=",")
```

## Descriptive Statistics, part 1:

First we will generate descriptive statistics for the groups of children being compared in the analysis of inflammation vs. no inflammation in NAFLD. 

```{r, echo=FALSE}
library(tableone)
listVars <- c("age", "sex", "race", "BMI", "ALT", "AST", "fibrosis_stage", 
              "lobular_inflammation", "portal_inflammation", "ballooning", "NAS_score")
catVars <- c("sex","race", "fibrosis_stage", "lobular_inflammation",
              "portal_inflammation", "ballooning", "NAS_score")
table1 <- CreateTableOne(vars = listVars, data = data, factorVars = catVars, 
                         strata=c("NASH_01_inflam"))
table1
```

## Descriptive Statistics, part 2:

Second we will generate descriptive statistics for the groups of children being compared in the analysis of NAS score < 3 versus 3 or greater among children with NAFLD. 

```{r, echo=FALSE}
library(tableone)
listVars <- c("age", "sex", "race", "BMI", "ALT", "AST", "fibrosis_stage",
              "lobular_inflammation", "portal_inflammation", "ballooning", "NAS_score")
catVars <- c("sex","race", "fibrosis_stage", "lobular_inflammation",
              "portal_inflammation", "ballooning", "NAS_score")
table1 <- CreateTableOne(vars = listVars, data = data, factorVars = catVars, 
                         strata=c("NASH_NAS"))
table1