---
title: "Descriptive Statistics - NAFLD vs. NASH Cohort"
author: "Catherine Cioffi"
date: "November 20, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction:

The goal of this analysis was to describe the metabolomic pathways associated with NAFLD disease progression. The following will generate descriptive Tables 1a-1c of the characteristics of the comparison groups for this analysis. The first comparison is based on fulfilling NASH criteria (yes/no). The second is based on presence of lobular or portal inflammation, versus no inflammation. The third comparison is based on a NAFLD Activity SCore (NAS) less than 3, versus greater than or equal to 3.

First we will read in the phenotype table from computer hard-drive:

```{r}
library(knitr)
library(kableExtra)
library(adm)
library(readxl)

#Option 1 - read in CSV from desktop. I will commment this out for now.
#data <- read.csv("C:/Users/cecioff/Desktop/metabolomics/_CHEAR/NASH_panel/phenotype_data.csv", header=TRUE, sep=",")

#Option 2 - read in dataset from working directory:
data <- read.csv("phenotype_data.csv", header=TRUE, na.strings=c("", "NA"), sep=",")

```

##SELECT VARIABLES & VIEW THE DATASET:

Below we will select the variables of interest from the phenotype dataset and display.

```{r}
#first select subjects we want, i.e. those with liver biopsy data and NAFLD:
data <- subset(data, (NAFLD_01 == 1 & is.na(NASH_criteria_yn) == FALSE & NASH_criteria_yn != "a"))

data_stripped <- data[, c("CHEAR_ID", "NAFLD_01", "NASH_01", "Inflam_01", "NAS_01", "T2DM", "IGT", "IFG", 'age', 'sex', 'race', 'BMI', 'BMIZ', 'BMIPCT', 'ALT', 'AST', 'Triglyc', 'HDL', 'LDL', 'Total_chol', 'fibrosis_stage', 'lobular_inflammation', 'portal_inflammation', 'ballooning', 'NAS_score')]

options(knitr.table.format = "html") 

data_stripped %>%
  kable("html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width=F, 
                position="left", font_size=10) %>%
  row_spec(0, angle=-45) %>%
  scroll_box(width = "750px", height = "750px")

```

## Descriptive Statistics, part 1:

First we will generate descriptive statistics for the groups of children being compared in the analysis of NASH vs. non-NASH in NAFLD. This is a simple way of doing this using R-package table one.

# Table 1:

We will take the information from above and input into formatted "Table 1".

```{r, results= "asis" }

# get n1 and n0;
data[,'NASH_01'] <- as.factor(data$NASH_01)
n <- summary(data$NASH_01)
n0str <- paste0("n=", n["0"])
n1str <- paste0("n=", n["1"])

# create the header
# note that &___; are called HTML entities and allow you produce special characters
header <- list()
HEADERATTR <- list(NULL, list(colspan = 2), NULL, list(colspan = 2), NULL, NULL)
header <- c(header, 
            html(
              html(
                # &nbsp; are non-breaking spaces for some padding
                # in the empty column
                c("", "NASH=yes", "&nbsp;&nbsp;&nbsp;", 
                  "NASH=no", "", ""), 
                "th", 
                HEADERATTR
              ), 
              "tr")
)

HEADERATTR2 <- mapply(c, 
                      HEADERATTR, 
                      list(NULL, list(style = "border-bottom: black solid"), NULL, 
                           list(style = "border-bottom: black solid"), NULL, NULL
                      )
)

header <- c(header, 
            html(
              html(
                c("", n1str, "", n0str, "", ""), 
                "th", 
                HEADERATTR2), 
              "tr")
)
header <- c(header, 
            html(
              html(
                c("Variable", "n", "%", "", "n", "%", 
                  "&chi;<sup>2</sup>", "p &dagger;"), 
                "th", 
                list(list(style = "border-bottom: black solid"))), 
              "tr")
)
header <- paste(header, collapse = "")

# make some derived variables
data$age.le12<- factor(ifelse(data$age <= 12, 1, 0))
data$age.13.15<- factor(ifelse(((data$age >= 13) & (data$age <= 15)), 1, 0))
data$age.ge16<- factor(ifelse(data$age >= 16, 1, 0))

data$sex.male<- factor(ifelse(data$sex == 1,1,0))
data$sex.female<- factor(ifelse(data$sex == 2,1,0))

data$White <- factor(ifelse(data$race == 2,1,0))
data$Black<- factor(ifelse(data$race == 1,1,0))
data$Hispanic <- factor(ifelse(data$race == 3,1,0))
data$Other<- factor(ifelse(data$race >= 4,1,0))

data$fibrosis.yes <-factor(ifelse(data$fibrosis_stage != 0, 1,0))
data$lobular.yes <-factor(ifelse(data$lobular_inflammation != 0, 1,0))
data$portal.yes <-factor(ifelse(data$portal_inflammation != 0, 1,0))
data$ballooning.yes <-factor(ifelse(data$ballooning == 1,1,0))

data$NAS_score <- as.numeric(data$NAS_score)
data$NAS.12 <-factor(ifelse(data$NAS_score < 3, 1,0))
data$NAS.34 <-factor(ifelse((data$NAS_score>=3 & data$NAS_score<5), 1,0))
data$NAS.ge5 <-factor(ifelse(data$NAS_score >= 5, 1,0))

row.names <- c(
  "Age",
  "&nbsp;-&nbsp; &le;12 yrs", 
  "&nbsp;-&nbsp; 13-15 yrs",
  "&nbsp;-&nbsp; &ge;16 yrs",
  "Male",
  "Female",
  "Race", 
  "&nbsp;-&nbsp; Black", 
  "&nbsp;-&nbsp; White",
  "&nbsp;-&nbsp; Hispanic",
  "&nbsp;-&nbsp; Other", 
  "Fibrosis, yes/no", 
  "Lobular Inflam, yes/no",
  "Portal inflam, yes/no",
  "Ballooning, yes/no",
  "NAS_score",
  "&nbsp;-&nbsp; 1-2", 
  "&nbsp;-&nbsp; 3-4",
  "&nbsp;-&nbsp; &ge;5"
)
  
# build a function to output each row
row.builder <- function(varname) {
  if(varname == "NULL") return(rep("", 7))
  tab <- table(data[, varname], data$NASH_01)
  props <- sprintf("%.1f", prop.table(tab, 2)[2, ]*100)
  stats <- summary(tab)
  # automatically determine if Fisher test is appropriate
  if(stats$approx.ok) {
    chisq <- sprintf("%.2f", stats$statistic)
    pval <- stats$p.value
    twotail <- ""
  } else {
    chisq <- "&Dagger;"
    pval <- fisher.test(tab)$p.value
    twotail <- "&sect;"
  }
  star <- ifelse(pval<0.05, "*", "")
  pval <- ifelse(pval<0.001, "&lt;0.001", 
                 ifelse(pval<0.01, sprintf("%.3f", stats$p.value), 
                        sprintf("%.2f", stats$p.value)))
  pval <- paste(pval, star, twotail, sep = "")
  c(tab[2, 2], props[2], "", tab[2, 1], props[1], chisq, pval)
}

vars <- c("NULL", "age.le12", "age.13.15", "age.ge16", "sex.male", "sex.female", "NULL", 
          "Black", "White", "Hispanic", "Other", "fibrosis.yes", "lobular.yes", "portal.yes", 
          "ballooning.yes", "NULL", "NAS.12", "NAS.34", "NAS.ge5")

body <- do.call(rbind, lapply(vars, row.builder))

# all but last row
row.names1 <- sapply(row.names[-length(row.names)], 
                     function(x) html(x, "td", list(list(style="font: bold"))))
body1 <- apply(body[-nrow(body), ], 1, 
               function(x) html(x, "td", list(list(style = "text-align: center"))))

# last row special formatting
row.names2 <- html(row.names[length(row.names)], "td", 
                   list(list(style = "font: bold; border-bottom: black solid")))
body2 <- html(body[nrow(body), ], "td", 
              list(list(style = "text-align: center; border-bottom: black solid")))

# recombine:
row.names <- c(row.names1, row.names2)
body <- c(body1, body2)
body <- html(paste(row.names, body), "tr")
all <- paste(header, body, sep = "")

# title:
cat(html("<b>Table 1.</b> Demographic, clinical, and histological charactericteristics of children with NAFLD with or without NASH", "p"), 
    file = "table1.html")

# table output
table <- html(all, "table", list(list(style = "border-collapse:collapse")))
cat(table, file = "table1.html", append = TRUE)

# footnotes
footnotes <- c(
  "&dagger; &alpha;=0.05. Significant p-values are marked with an asterisk (*).<br/>", 
  "&Dagger; Fisher's Exact Test<br/>", 
  "&sect; Two-tailed p-value<br/>")
cat(html(footnotes, "small"), file = "table1.html", append = TRUE)

# file was written above but we'll read it back in and display here
cat(readLines("table1.html"))

```

## Descriptive Statistics, part 2:

First we will generate descriptive statistics for the groups of children being compared in the analysis of inflammation vs. no inflammation in NAFLD. This is a simple way of doing this using R-package table one.

# Table 1b:

We will take the information from above and input into formatted "Table 1b".

```{r, results= "asis" }

# get n1 and n0;
data[,'NASH_inflam'] <- as.factor(data$Inflam_01)
n <- summary(data$NASH_inflam)
n0str <- paste0("n=", n["0"])
n1str <- paste0("n=", n["1"])

# create the header
# note that &___; are called HTML entities and allow you produce special characters
header <- list()
HEADERATTR <- list(NULL, list(colspan = 2), NULL, list(colspan = 2), NULL, NULL)
header <- c(header, 
            html(
              html(
                # &nbsp; are non-breaking spaces for some padding
                # in the empty column
                c("", "Inflammation", "&nbsp;&nbsp;&nbsp;", 
                  "No Inflammation", "", ""), 
                "th", 
                HEADERATTR
              ), 
              "tr")
)

HEADERATTR2 <- mapply(c, 
                      HEADERATTR, 
                      list(NULL, list(style = "border-bottom: black solid"), NULL, 
                           list(style = "border-bottom: black solid"), NULL, NULL
                      )
)

header <- c(header, 
            html(
              html(
                c("", n1str, "", n0str, "", ""), 
                "th", 
                HEADERATTR2), 
              "tr")
)
header <- c(header, 
            html(
              html(
                c("Variable", "n", "%", "", "n", "%", 
                  "&chi;<sup>2</sup>", "p &dagger;"), 
                "th", 
                list(list(style = "border-bottom: black solid"))), 
              "tr")
)
header <- paste(header, collapse = "")

# make some derived variables
data$age.le12<- factor(ifelse(data$age <= 12, 1, 0))
data$age.13.15<- factor(ifelse(((data$age >= 13) & (data$age <= 15)), 1, 0))
data$age.ge16<- factor(ifelse(data$age >= 16, 1, 0))

data$sex.male<- factor(ifelse(data$sex == 1,1,0))
data$sex.female<- factor(ifelse(data$sex == 2,1,0))

data$White <- factor(ifelse(data$race == 2,1,0))
data$Black<- factor(ifelse(data$race == 1,1,0))
data$Hispanic <- factor(ifelse(data$race == 3,1,0))
data$Other<- factor(ifelse(data$race >= 4,1,0))

data$fibrosis.yes <-factor(ifelse(data$fibrosis_stage != 0, 1,0))
data$lobular.yes <-factor(ifelse(data$lobular_inflammation != 0, 1,0))
data$portal.yes <-factor(ifelse(data$portal_inflammation != 0, 1,0))
data$ballooning.yes <-factor(ifelse(data$ballooning == 1,1,0))

data$NAS_score <- as.numeric(data$NAS_score)
data$NAS.12 <-factor(ifelse(data$NAS_score < 3, 1,0))
data$NAS.34 <-factor(ifelse((data$NAS_score>=3 & data$NAS_score<5), 1,0))
data$NAS.ge5 <-factor(ifelse(data$NAS_score >= 5, 1,0))

row.names <- c(
  "Age",
  "&nbsp;-&nbsp; &le;12 yrs", 
  "&nbsp;-&nbsp; 13-15 yrs",
  "&nbsp;-&nbsp; &ge;16 yrs",
  "Male",
  "Female",
  "Race", 
  "&nbsp;-&nbsp; Black", 
  "&nbsp;-&nbsp; White",
  "&nbsp;-&nbsp; Hispanic",
  "&nbsp;-&nbsp; Other", 
  "Fibrosis, yes/no", 
  "Lobular Inflam, yes/no",
  "Portal inflam, yes/no",
  "Ballooning, yes/no",
  "NAS_score",
  "&nbsp;-&nbsp; 1-2", 
  "&nbsp;-&nbsp; 3-4",
  "&nbsp;-&nbsp; &ge;5"
)
  
# build a function to output each row
row.builder <- function(varname) {
  if(varname == "NULL") return(rep("", 7))
  tab <- table(data[, varname], data$NASH_inflam)
  props <- sprintf("%.1f", prop.table(tab, 2)[2, ]*100)
  stats <- summary(tab)
  # automatically determine if Fisher test is appropriate
  if(stats$approx.ok) {
    chisq <- sprintf("%.2f", stats$statistic)
    pval <- stats$p.value
    twotail <- ""
  } else {
    chisq <- "&Dagger;"
    pval <- fisher.test(tab)$p.value
    twotail <- "&sect;"
  }
  star <- ifelse(pval<0.05, "*", "")
  pval <- ifelse(pval<0.001, "&lt;0.001", 
                 ifelse(pval<0.01, sprintf("%.3f", stats$p.value), 
                        sprintf("%.2f", stats$p.value)))
  pval <- paste(pval, star, twotail, sep = "")
  c(tab[2, 2], props[2], "", tab[2, 1], props[1], chisq, pval)
}

vars <- c("NULL", "age.le12", "age.13.15", "age.ge16", "sex.male", "sex.female", "NULL", 
          "Black", "White", "Hispanic", "Other", "fibrosis.yes", "lobular.yes", "portal.yes", 
          "ballooning.yes", "NULL", "NAS.12", "NAS.34", "NAS.ge5")

body <- do.call(rbind, lapply(vars, row.builder))

# all but last row
row.names1 <- sapply(row.names[-length(row.names)], 
                     function(x) html(x, "td", list(list(style="font: bold"))))
body1 <- apply(body[-nrow(body), ], 1, 
               function(x) html(x, "td", list(list(style = "text-align: center"))))

# last row special formatting
row.names2 <- html(row.names[length(row.names)], "td", 
                   list(list(style = "font: bold; border-bottom: black solid")))
body2 <- html(body[nrow(body), ], "td", 
              list(list(style = "text-align: center; border-bottom: black solid")))

# recombine:
row.names <- c(row.names1, row.names2)
body <- c(body1, body2)
body <- html(paste(row.names, body), "tr")
all <- paste(header, body, sep = "")

# title:
cat(html("<b>Table 1b.</b> Inflammation vs. No-Inflammation among Children with NAFLD", "p"), 
    file = "table1b.html")

# table output
table <- html(all, "table", list(list(style = "border-collapse:collapse")))
cat(table, file = "table1b.html", append = TRUE)

# footnotes
footnotes <- c(
  "&dagger; &alpha;=0.05. Significant p-values are marked with an asterisk (*).<br/>", 
  "&Dagger; Fisher's Exact Test<br/>", 
  "&sect; Two-tailed p-value<br/>")
cat(html(footnotes, "small"), file = "table1b.html", append = TRUE)

# file was written above but we'll read it back in and display here
cat(readLines("table1b.html"))

```

## Descriptive Statistics, part 3:

Second we will generate descriptive statistics for the groups of children being compared in the analysis of NAS score < 3 versus 3 or greater among children with NAFLD. Again, here is a simple way of doing this.

# Table 1c:

We will take the information from above and input into formatted "Table 1c".

```{r, results="asis"}
# get n1 and n0;
data[,'NAS_01'] <- as.factor(data$NAS_01)
n <- summary(data$NAS_01)
n0str <- paste0("n=", n["0"])
n1str <- paste0("n=", n["1"])

# create the header
# note that &___; are called HTML entities and allow you produce special characters
header <- list()
HEADERATTR <- list(NULL, list(colspan = 2), NULL, list(colspan = 2), NULL, NULL)
header <- c(header, 
            html(
              html(
                # &nbsp; are non-breaking spaces for some padding
                # in the empty column
                c("", "NAS Score &ge;3", "&nbsp;&nbsp;&nbsp;", 
                  "NAS Score &lt;3", "", ""), 
                "th", 
                HEADERATTR
              ), 
              "tr")
)

HEADERATTR2 <- mapply(c, 
                      HEADERATTR, 
                      list(NULL, list(style = "border-bottom: black solid"), NULL, 
                           list(style = "border-bottom: black solid"), NULL, NULL
                      )
)

header <- c(header, 
            html(
              html(
                c("", n1str, "", n0str, "", ""), 
                "th", 
                HEADERATTR2), 
              "tr")
)
header <- c(header, 
            html(
              html(
                c("Variable", "n", "%", "", "n", "%", 
                  "&chi;<sup>2</sup>", "p &dagger;"), 
                "th", 
                list(list(style = "border-bottom: black solid"))), 
              "tr")
)
header <- paste(header, collapse = "")

# make some derived variables
data$age.le12<- factor(ifelse(data$age <= 12, 1, 0))
data$age.13.15<- factor(ifelse(((data$age >= 13) & (data$age <= 15)), 1, 0))
data$age.ge16<- factor(ifelse(data$age >= 16, 1, 0))

data$sex.male<- factor(ifelse(data$sex == 1,1,0))
data$sex.female<- factor(ifelse(data$sex == 2,1,0))

data$White <- factor(ifelse(data$race == 2,1,0))
data$Black<- factor(ifelse(data$race == 1,1,0))
data$Hispanic <- factor(ifelse(data$race == 3,1,0))
data$Other<- factor(ifelse(data$race >= 4,1,0))

data$fibrosis.yes <-factor(ifelse(data$fibrosis_stage != 0, 1,0))
data$lobular.yes <-factor(ifelse(data$lobular_inflammation != 0, 1,0))
data$portal.yes <-factor(ifelse(data$portal_inflammation != 0, 1,0))
data$ballooning.yes <-factor(ifelse(data$ballooning == 1,1,0))

data$NAS_score <- as.numeric(data$NAS_score)
data$NAS.12 <-factor(ifelse(data$NAS_score < 3, 1,0))
data$NAS.34 <-factor(ifelse((data$NAS_score>=3 & data$NAS_score<5), 1,0))
data$NAS.ge5 <-factor(ifelse(data$NAS_score >= 5, 1,0))

row.names <- c(
  "Age",
  "&nbsp;-&nbsp; &le;12 yrs", 
  "&nbsp;-&nbsp; 13-15 yrs",
  "&nbsp;-&nbsp; &ge;16 yrs",
  "Male",
  "Female",
  "Race", 
  "&nbsp;-&nbsp; Black", 
  "&nbsp;-&nbsp; White",
  "&nbsp;-&nbsp; Hispanic",
  "&nbsp;-&nbsp; Other", 
  "Fibrosis, yes/no", 
  "Lobular Inflam, yes/no",
  "Portal inflam, yes/no",
  "Ballooning, yes/no",
  "NAS_score",
  "&nbsp;-&nbsp; 1-2", 
  "&nbsp;-&nbsp; 3-4",
  "&nbsp;-&nbsp; &ge;5"
)
  
# build a function to output each row
row.builder <- function(varname) {
  if(varname == "NULL") return(rep("", 7))
  tab <- table(data[, varname], data$NAS_01)
  props <- sprintf("%.1f", prop.table(tab, 2)[2, ]*100)
  stats <- summary(tab)
  # automatically determine if Fisher test is appropriate
  if(stats$approx.ok) {
    chisq <- sprintf("%.2f", stats$statistic)
    pval <- stats$p.value
    twotail <- ""
  } else {
    chisq <- "&Dagger;"
    pval <- fisher.test(tab)$p.value
    twotail <- "&sect;"
  }
  star <- ifelse(pval<0.05, "*", "")
  pval <- ifelse(pval<0.001, "&lt;0.001", 
                 ifelse(pval<0.01, sprintf("%.3f", stats$p.value), 
                        sprintf("%.2f", stats$p.value)))
  pval <- paste(pval, star, twotail, sep = "")
  c(tab[2, 2], props[2], "", tab[2, 1], props[1], chisq, pval)
}

vars <- c("NULL", "age.le12", "age.13.15", "age.ge16", "sex.male", "sex.female", "NULL", 
          "Black", "White", "Hispanic", "Other", "fibrosis.yes", "lobular.yes", "portal.yes", 
          "ballooning.yes", "NULL", "NAS.12", "NAS.34", "NAS.ge5")

body <- do.call(rbind, lapply(vars, row.builder))

# all but last row
row.names1 <- sapply(row.names[-length(row.names)], 
                     function(x) html(x, "td", list(list(style="font: bold"))))
body1 <- apply(body[-nrow(body), ], 1, 
               function(x) html(x, "td", list(list(style = "text-align: center"))))

# last row special formatting
row.names2 <- html(row.names[length(row.names)], "td", 
                   list(list(style = "font: bold; border-bottom: black solid")))
body2 <- html(body[nrow(body), ], "td", 
              list(list(style = "text-align: center; border-bottom: black solid")))

# recombine:
row.names <- c(row.names1, row.names2)
body <- c(body1, body2)
body <- html(paste(row.names, body), "tr")
all <- paste(header, body, sep = "")

# title:
cat(html("<b>Table 1c.</b> NAS Score &lt;3 vs. NAS Score &ge;3 among Children with NAFLD", "p"), 
    file = "table1c.html")

# table output
table <- html(all, "table", list(list(style = "border-collapse:collapse")))
cat(table, file = "table1c.html", append = TRUE)

# footnotes
footnotes <- c(
  "&dagger; &alpha;=0.05. Significant p-values are marked with an asterisk (*).<br/>", 
  "&Dagger; Fisher's Exact Test<br/>", 
  "&sect; Two-tailed p-value<br/>")
cat(html(footnotes, "small"), file = "table1c.html", append = TRUE)

# file was written above but we'll read it back in and display here
cat(readLines("table1c.html"))
```